---
title: "Bee microbiomes in a changing climate"
author: ""
date: ""
output: html_document
---
#### **Purpose:** Analysis of treatment, life history & health data

#### **Ownership:** Bailey Crowley & Robert Schaeffer


#### Prepare workspace


```{r, warning = FALSE}
# Set working directory
  setwd("~/Downloads")

# Import data
  all_sensors <- read.csv("Research/OsmiaCC/Monnit_sensors/all_sensors.csv")
  health <- read.csv("health - Raw Data.csv")
  duration <- read.csv("life_history - Data.csv")
  mortality <- read.csv("mortality - Data.csv")
```


```{r, message = FALSE}
# Load necessary packages
  library(tidyverse)
  library(lubridate)
  library(dplyr)
  library(readr)
  library(xts)
  library(ggplot2)
  library(scales)
  library(multcompView)
  library(gridExtra)
  library(cowplot)
  library(knitr)
  library(survival)
  library(tibble)
  library(ggsurvfit)
```


#### Analysis of temperature treatment data


```{r}
# Remove dates after July 4, when the experiment stopped
  all_sensors <- all_sensors %>% filter(Date < '2023-07-04')

# Format date column
  all_sensors$Date <- as.Date(all_sensors$Date, format = "%Y-%m-%d")
  
# Combine date and time columns
  all_sensors$DateTime <- as.POSIXct(paste(all_sensors$Date, all_sensors$Time),
                                     format = "%Y-%m-%d %I:%M:%S %p")
```


```{r}
# Manually order legend
  all_sensors$Sensor <- factor(all_sensors$Sensor, levels = c("Warm", "Ambient", "Cool"))
```


```{r, fig.height = 8, fig.width = 18}
# Plot temperature & humidity
  ggplot(all_sensors, aes(x = DateTime, colour = Sensor, group = Sensor)) +
              geom_line(aes(y = Temp, 
                            linetype = "Temperature")) + 
              geom_line(aes(y = Humidity, 
                            linetype = "Humidity")) +
              theme_bw() +
              theme(legend.title = element_blank()) +
              theme(text = element_text(size = 16)) +
              xlab("Date") +
              scale_y_continuous(name = "Temperature (°C)", 
                                 sec.axis = sec_axis(trans = ~.*1, name = "Relative Humidity (%)")) +
              scale_color_manual(values = c("#C62828",  "#616161", "#1565C0"))
```


#### Clean data



```{r}
# Remove rows with no data
  health <- na.omit(health)

# Subset to create dfs by sex
  males_health <- health[health$sex == "M", ]
  males_duration <- duration[duration$sex == "M", ]
  males_mortality <- mortality[mortality$sex == "M", ]
```


#### Larval body mass


```{r}
# Determine sample sizes by sex & treatment
  males_health_ss <- males_health %>%
    group_by(combo_treat) %>%
    tally()
  males_health_ss
```


```{r}
# Subset df to include just treatments and response variable (wet body mass)
  males_health_mass <- males_health %>%
    select(bee, temp_treat, micro_treat, combo_treat, wet_mass_mg)
```


```{r}
# Do any of the group means differ?  
  males_mass_lm <- lm(wet_mass_mg ~ temp_treat * micro_treat, data = males_health_mass)
  males_mass_aov <- aov(males_mass_lm)
  summary(males_mass_aov)
```


INTERPRETATION HERE


```{r}
# Which group means differ?  
  mass_tukey <- TukeyHSD(males_mass_aov)
  mass_tukey
```


INTERPRETATION HERE


```{r}
# Reorder the x-axis
  males_health_mass$combo_treat <- factor(males_health_mass$combo_treat, levels = c("CS", "CN", "AS", "AN", "WS", "WN"))
```


```{r}
# Plot larval body mass by treatment
ggplot(males_health_mass, aes(x = combo_treat, y = wet_mass_mg,color = combo_treat)) + 
    geom_boxplot(outlier.shape = NA, 
                 width = 0.5, 
                 position = position_dodge(width = 0.1)) + 
    geom_jitter(size = 1, 
                alpha = 0.9) +
    theme_bw() +
    theme(text = element_text(size = 16)) +
    ylim(0, 20) +
    scale_fill_discrete(labels = c("CS", "CN", "AS", "AN", "WS", "WN")) +
    scale_x_discrete(labels = c("CS", "CN", "AS", "AN", "WS", "WN")) + 
    scale_color_manual(name = "Treatment", 
                       values = c("#64B5F6","#1565C0", "#9E9E9E", "#616161", "#E57373", "#C62828"),
                       labels = c('Cool: Sterile', 'Cool: Natural', 'Ambient: Sterile', 'Ambient: Natural', 'Warm: Sterile', 'Warm: Natural')) +
    ggtitle("") +
    ylab("Larval body mass (mg)") + 
    xlab("Treatments")
```


INTERPRETATION


#### Proportion of larval body fat


```{r}
# Subset df to include just treatments and response variable (wet body mass)
  males_health_fat <- males_health %>%
    select(bee, temp_treat, micro_treat, combo_treat, prop_body_fat)
```


```{r}
# Do any of the group means differ?
  males_fat_lm <- lm(prop_body_fat ~ temp_treat * micro_treat, data = males_health_fat)
  males_fat_aov <- aov(males_fat_lm)
  summary(males_fat_aov)
```


```{r}
# Which group means differ?
  fat_tukey <- TukeyHSD(males_fat_aov)
  fat_tukey
```


INTERPRETATION HERE


```{r}
# Reorder the x-axis
  males_health_fat$combo_treat <- factor(males_health$combo_treat, levels = c("CS", "CN", "AS", "AN", "WS", "WN"))
```


```{r}
# Plot the proportion of larval body fat by treatment
ggplot(males_health_fat, aes(x = combo_treat, y = prop_body_fat, color = combo_treat)) + 
              geom_boxplot(outlier.shape = NA, 
                           width = 0.5, 
                           position = position_dodge(width = 0.1)) + 
              geom_jitter(size = 1, 
                          alpha = 0.9) +
              theme_bw() +
              theme(text = element_text(size = 16)) +
              ylim(0, 6) +
              scale_x_discrete(labels = c("CS", "CN", "AS", "AN", "WS", "WN")) + 
              scale_color_manual(name = "Treatment", 
                                 values = c("#64B5F6","#1565C0", "#9E9E9E", "#616161", "#E57373", "#C62828"),
                                 labels = c('Cool: Sterile', 'Cool: Natural', 'Ambient: Sterile', 'Ambient: Natural', 'Warm: Sterile', 'Warm: Natural')) +
              ggtitle("") +
              ylab("Proportion of body fat") + 
              xlab("Treatment")
```


INTERPRETATION HERE


#### Duration of developmental stages


```{r}
# Group males by treatment, then find the mean and sd for instars 1 & 2
  males_duration_table <- males_duration %>%
    group_by(combo_treat) %>%
    summarise(mean_instar1 = mean(days_instar1.2),
              sd_instar1 = sd(days_instar1.2),
              mean_instar2 = mean(days_instar2.5),
              sd_instar2 = sd(days_instar2.5))
  males_duration_table
```


```{r}
# Do any of the group means differ?
  males_dev_lm <- lm(days_instar2.5 ~ temp_treat*micro_treat, data = males_duration)
  anova_dev <- aov(males_dev_lm)
  summary(anova_dev)
```


INTERPRETATION HERE


```{r}
# Which group means differ?
  ph_dev_tukey <- TukeyHSD(anova_dev)
  ph_dev_tukey
```


INTERPRETATION HERE


```{r}
# Reorder the x-axis
  males_health$combo_treat <- factor(males_health$combo_treat, levels = c("CS", "CN", "AS", "AN", "WS", "WN"))
```


```{r, warning = FALSE, message = FALSE}
# Plot larval duration by treatment
  ggplot(males_duration, aes(x = combo_treat, y = days_instar2.5, color = combo_treat)) + 
        geom_boxplot(outlier.shape = NA) + 
        geom_jitter(size = 1, 
                    alpha = 0.9) +
        theme_bw() +
        theme(text = element_text(size = 16)) +
        ylim(5, 20) +
        scale_fill_discrete(labels = c("CS", "CN", "AS", "AN", "WS", "WN")) +
        scale_x_discrete(labels = c("CS", "CN", "AS", "AN", "WS", "WN")) + 
        scale_color_manual(name = "Treatment", 
                           values = c("#64B5F6","#1565C0", "#9E9E9E", "#616161", "#E57373", "#C62828"),
                           labels = c('Cool: Sterile', 'Cool: Natural', 'Ambient: Sterile', 'Ambient: Natural', 'Warm: Sterile', 'Warm: Natural')) +             
              ggtitle("") +
              ylab("Duration Larval instars II-V (days)") + 
              xlab("Treatment")
```


INTERPRETATION HERE


#### Mortality


```{r}
# Table of sample size per treatment
  males_mortality_ss <- males_mortality %>%
    group_by(combo_treat) %>%
    tally()
  males_mortality_ss
```


#### Survivorship analysis


Helpful resource: https://www.emilyzabor.com/tutorials/survival_analysis_in_r_tutorial.html


Status: 0 = survival to the fifth instar; 1 = death


```{r, warning = FALSE}
# Convert chr to date
  males_duration <- males_duration %>%
                          mutate(
                            date_nesting_start = ymd(date_nesting_start),
                            date_nesting_end = ymd(date_nesting_end),
                            date_graft = ymd(date_graft),
                            date_instar1 = ymd(date_instar1),
                            date_instar2 = ymd(date_instar2),
                            date_instar5 = ymd(date_instar5),
                            date_last_alive = ymd(date_last_alive)
                          )
```


```{r}
# Check to see if it worked  
  tibble(males_duration)
```


```{r}
# Calculate the number of days bees survived
   males_duration <- males_duration %>%
                          mutate(
                            total_surv_days = as.duration(date_graft %--% date_last_alive / ddays(1))
                          )
```


```{r}
# Format chr to numeric
   males_duration$total_surv_days <- as.numeric(males_duration$total_surv_days)
```


```{r}
# Check to see if it worked   
   head(males_duration)
```


```{r}
# Create a survival object
   Surv(males_duration$total_surv_days, males_duration$status)
```


```{r}
# Create a survival curve
  survfit(Surv(total_surv_days, status) ~ combo_treat, data = males_duration)
```


```{r}
# Fit the line
  s2 <- survfit2(Surv(total_surv_days, status) ~ combo_treat, data = males_duration)
```


```{r}
# Plot Kaplan-Meier
  ggsurvfit(s2) +
      theme_classic() +
      theme(legend.position = "right") +
      theme(text = element_text(size = 16)) +
      scale_color_manual(name = "Treatment", 
                         values = c("#64B5F6","#1565C0", "#9E9E9E", "#616161", "#E57373", "#C62828"),
                         labels = c('Cool: Sterile', 'Cool: Natural', 'Ambient: Sterile', 'Ambient: Natural', 'Warm: Sterile', 'Warm: Natural')) +
      scale_y_continuous(limits = c(0, 1)) +
      scale_x_continuous(breaks = c(0, 5, 10, 15, 20, 25)) +
      labs(x = "Days", y = "Survival probability")
```


INTERPRETATION


```{r}
# Cox regression model
  coxph(Surv(total_surv_days, status) ~ combo_treat, data = males_duration)
```


INTERPRETATION HERE


```{r}
# Compare survival times between treatments
  survdiff(Surv(total_surv_days, status) ~ combo_treat, data = males_duration)
```

